Inferring stellar rotation periods using Gaussian processes

### 2016-12-06 documents/ms.tex
lots of changes

\newcommand{\columbia}{1}
\newcommand{\oxford}{2}
\newcommand{\uw}{3}
\newcommand{\princeton}{4}
\newcommand{\nlightcurves}{333}
\newcommand{\Kepler}{{\it Kepler}}
\newcommand{\kepler}{\Kepler}
\newcommand{\corot}{{\it CoRoT}}
\newcommand{\Ktwo}{{\it K2}}
\newcommand{\ktwo}{\Ktwo}
\newcommand{\TESS}{{\it TESS}}
\newcommand{\LSST}{{\it LSST}}
\newcommand{\SDSS}{{\it SDSS}}
\newcommand{\PLATO}{{\it PLATO}}
\newcommand{\Teff}{$T_{\mathrm{eff}}$}
\newcommand{\teff}{$T_{\mathrm{eff}}$}
\newcommand{\FeH}{[Fe/H]}
\newcommand{\feh}{[Fe/H]}
\newcommand{\ie}{{\it i.e.}}
\newcommand{\eg}{{\it e.g.}}
\newcommand{\logg}{log \emph{g}}
\newcommand{\dnu}{$\Delta \nu$}
\newcommand{\numax}{$\nu_{\mathrm{max}}$}

@@

\altaffiltext{\columbia}{Simons Fellow, Columbia University, NY, NY,
ruthangus@gmail.com}
\altaffiltext{\oxford}{Subdepartment of Astrophysics, University of Oxford,
OX1 3RH, UK}
\altaffiltext{\uw}{Department of Astronomy, University of Washington, Seattle,
WA}
\altaffiltext{\princeton}{Department of Astronomy, Princeton University}

\begin{abstract}
The light curves of spotted, rotating stars are often non-sinusoidal and
quasi-periodic---spots are not static on the stellar surface and have finite
lifetimes causing stellar flux variations to slowly shift in phase.
A strictly periodic sinusoid is therefore not a representative model for the
light curve of star with a rotational signal.
Ideally, a physical model of the stellar surface would be conditioned on the
data, however the parameters of such models can be highly degenerate
\citep[\eg][]{Russell1906, Jeffers2009}.
Instead, we use an appropriate {\it effective} model: a Gaussian Process (GP)
with a quasi-periodic covariance kernel function.
By modelling the light curve with a GP, a highly flexible semi-parametric
function, we avoid the necessity of choosing a `best fitting' functional form,
whilst sampling directly from the posterior probability distribution function
of the periodic parameter and marginalising over the other kernel
hyperparameters.
We used \nlightcurves\ simulated light curves with a range of rotation periods
to test the GP model.
We attempted to recover the rotation periods using three methods: our GP
method, a sine-fitting periodogram method and an autocorrelation function
method.
The posterior probability distribution of the rotation period parameter was
sampled using the affine invariant ensemble Markov Chain Monte Carlo (MCMC)
sampler {\tt emcee} \citep{Foreman-Mackey2013}, and the GP operations were
performed using the {\tt george} python package \citep{George}.
Rotation periods inferred via this method are more precise and accurate than
the periodogram and ACF methods.
Furthermore, the improvement is expected to be even more dramatic when applied
to real, noisy {\it Kepler} light curves, since the GP method is well suited
to modelling rotation signals and correlated noise simultaneously.

@@

In order to test the GP method we attempted to the measure rotation periods of
a set of simulated light curves.
We used \nlightcurves\ light curves generated for the \citet{Aigrain2015}
`hare and hounds' rotation period recovery experiment.
These light curves were simulated by placing dark, circular spots with slowly
evolving size, on the surface of bright, rotating spheres, ignoring
limb-darkening effects.
\citet{Aigrain2015} simulated one thousand light curves in order to test the
ability of participating teams to recover both the stellar rotation periods
and the rotational shear: the amplitude of surface differential rotation.
% Since we are not interested in recovering differential rotation we only used
% those light curves simulated {\it without} differential rotation, of which
% there are \nlightcurves.
For the purpose of demonstrating that our method can recover rotation periods,
we selected only the \nlightcurves\ out of the 1000 without differential
rotation.
We opted to use only solid-body rotators because differential rotation may
produce some additional scatter in the rotation period measurements recovered.
In future we intend to test whether we can recover differential rotation using
the GP method and will then use the full set of 1000 light curves.
The distributions of input parameters used to generate these light curves are
listed in \citet{Aigrain2015}.
Light curves were simulated with a real \Kepler\ long-cadence time array:
one data point every thirty minutes over a four year duration.
90\% of the rotation periods of the simulations were randomly drawn from a
log-uniform distribution between 10 and 50 days and 10\% from a log-uniform
distribution between 1 and 10 days.
The light curves were also generated with a range of stellar inclination
angles, activity levels, spot lifetimes and more.
Once simulated, the light curves were added to real \kepler\ light curves with
no obvious astrophysical variability in order to preserve the noise properties
of the time series.
Figure \ref{fig:demo_lc} shows an example of a simulated, noise-free
light curve with a period of 17.04 days.

@@

\begin{table*}
\begin{center}
\caption{Ranges and distributions of parameters used to simulate light curves
in \citet{aigrain2015}}
\begin{tabular}{lcc}
\hline\hline
    Parameter & Range & Distribution \\
    \hline
    Rotation period, $P_{rot}$ & 10 - 50 days (90\%) & log uniform \\
    & 1 - 10 days (10\%) & log uniform \\
    Activity cycle length & 1 - 10 years & log uniform \\
    Inclination & 0 - 90$^\circ$ & Uniform in $\sin^2i$ \\
    Decay timescale & (1 - 10) $\times P_{rot}$ & log uniform \\
\hline
\end{tabular}
\end{center}
\end{table*}
\label{tab:simulation_parameters}

@@

We attempted to recover the rotation periods of these \nlightcurves\ light
curves using three rotation period recovery methods: the ACF method, the LS
periodogram method and the GP method.

@@

We applied the ACF method to the sample of \nlightcurves\ simulated light
curves.
Periods measured using the ACF method are plotted against the original
rotation period values used to generate the light curves in figure
\ref{fig:compare_acf}.
The $2\times$ and $1/2\times$ harmonic lines are shown as dashed lines.
The agreement between the injected and recovered rotation periods is good in
general, however twice the true rotation period is often found.
In addition, some rotation periods are drastically over- and underestimated.

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/compare_acf.png}
\caption[ACF results.]
{The `true' rotation periods used to generate \nlightcurves\ simulated light
curves vs the rotation periods measured using the ACF technique.}
\label{fig:compare_acf}
\end{center}
\end{figure*}

@@

For each simulated light curve, a Lomb-Scargle (LS) periodogram
\citep{Lomb1976, Scargle1982} was computed over a grid of 10,000 periods,
evenly spaced in frequency, between 1 and 100 days.
The period of the highest peak in the periodogram was adopted as the rotation
period.
The resulting recovered rotation periods are plotted as a function of the true
periods in figure \ref{fig:pgram_compare}.
These recovered rotation periods are in general more accurate than the ACF
results: they do not systematically over or under predict rotation period.
They are however less precise than the ACF results---the 8.03 day RMS of
residuals is five times larger.
The longer-period measurements provide the dominant contribution to this RMS.
% We believe that the unsuccessful recoveries occurring preferentially for
% shorter rotation periods are produced by additional, longer timescale
% variations in some of the light curves.
% Specifically, by fluctuations produced by changes in the total spot coverage
% on the stellar surface.
% These fluctuations can occasionally be larger in magnitude than those produced
% by the rotational signal itself and are present in the majority of outlying
% light curves that occupy the top left of figure
% \ref{fig:pgram_compare_noise-free}.

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/compare_pgram.png}
\caption[LS periodogram results.]
{The `true' rotation periods used to generate \nlightcurves\ simulated light
curves vs the rotation periods measured using a LS periodogram technique.}
\label{fig:pgram_compare}
\end{center}
\end{figure*}

@@

\begin{table*}
\begin{center}
\caption{Means and standard deviations of the Gaussian priors on the natural
logarithms of the GP model parameters.}
\begin{tabular}{lcc}
Parameter & Mean & Standard Deviation \\
    \hline
    $A$ & -13 & 2.7 \\
    $l$ & 6.2 & 1.5 \\
    $\Gamma$ & -1.4 & 1.5 \\
    Rotation period, $P_{rot}$ & ACF period & ACF period $\times 2$ \\
    $\sigma$ & -17 & 5 \\
\end{tabular}
\end{center}
\end{table*}
\label{tab:priors}
The parameter positions were also initialised at the means of these
distributions during MCMC posterior exploration.

In addition, we set the following bounds on the rotation period and the $l$
parameter.
\begin{eqnarray}
 	&	\ln(0.5) < \ln(P_{\mathrm{rot}}) < \ln(T/2) \\ \nonumber
    &   0 < \ln(l) \\
\end{eqnarray}
\label{eq:priors}
Where $T$ is the total timespan of the light curve.
We limit rotation periods to be greater than half a day because we are chiefly
interested in rotation of FGK stars which are likely to be greater than this
lower limit, however this restriction could be relaxed in principle.
We also limit the rotation periods to be less than half of the total timespan
of the data since this is the longest period we expect to reliable recover.
We prevent $\ln(l)$ from falling beneath zero both because we expect that
astrophysical signals in light curves should have longer correlation
timescales than this and in practice we found that if $l$ is unrestricted, it
can occasionally wander towards a small value, resulting in an incorrect
rotation period estimate.

@@

We subsampled the light curves in order to reduce computation time.
We randomly subsampled the data, retaining an average of four data points per
day.
This allowed us to recover periods down to half a day --- the minimum
timescale of interest for this project, whilst reducing computational costs.
The {\tt george} \citep{George} python package was used to implement the GP
model which uses the fast matrix solver, HODLR \citep{Ambikasaran2014}.
Matrix operations performed with HODLR scale as $N \log ^2 (N)$, where $N$ is
the number of data points.
We used {\tt emcee} \citep{Foreman-Mackey2013}, an affine invariant ensemble
sampler to explore the posterior PDFs of the model parameters.
The rotation period was taken as the sample with the maximum likelihood value.
The resulting measured rotation periods are compared to the true rotation
periods in figure \ref{fig:compare_mcmc}.

The marginal posterior distributions of the QP kernel hyperparameters for an
example noise-free light curve  with a 14.4 day rotation period, are shown in
figure \ref{fig:gp_posteriors}.
The light curve itself with the best fit GP model is shown in figure
\ref{fig:demo_lc_GP}.

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/demo_lc_GP.pdf}
\caption{A simulated light curve with a rotation period of 14.4 days.
The pink curve shows the GP model with the best-fit parameters.
The marginal posteriors of the GP hyper-parameters used to produce this fit
are shown in figure \ref{fig:gp_posteriors}.}
\label{fig:demo_lc_GP}
\end{center}
\end{figure*}

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/demo_triangle.pdf}
\caption{Marginal posterior PDFs of the QP GP model parameters. $\sigma$ is an
additional white noise term added to the diagonal elements of the covariance
matrix. It is the fraction by which the observational uncertainties have been
underestimated. If the errorbars reported on the data are too small, this
parameter will be non-zero. Since this light curve was simulated without
noise, this parameter is very close to zero.
The true rotation period of this star is 14.4 days.}
\label{fig:gp_posteriors}
\end{center}
\end{figure*}

@@

The noise-free light curves were injected into real \kepler\ light curves in
order to test the performance of the GP method in a realistic case.
Ideally, simulated light curves would have been injected into the \kepler\
pixel-level data, then the same detrending method that is applied to the
\kepler\ light curves would be applied to these light curves before attempting
to measure rotation periods for these stars.
However, since this work is designed to be more of a demonstration of the
overall efficacy of the GP periodogram, rather than the development of a
\Kepler\ specific method, this is beyond the scope of our demonstrations.

Unfortunately, automating the GP method so that it can run on large numbers of
real \kepler\ light curves is difficult.
The added noise reduces the acceptance fraction within the MCMC chains,
significantly increasing the required time for convergence.
Despite this, the method works well on individual targets where chains can be
run for a long time and subsampling can be reduced.
We used a simulated, noisy light curve from \citet{Aigrain2015} that was
generated by injecting a noise-free simulation into a real \kepler\ light
curve in order to preserve realistic \kepler\ noise properties.

\kepler\ light curves are big data and extracting the maximum amount of
information from them requires either large numbers of CPU hours, or using
work-arounds.
One feature of \kepler\ light curves that works to our advantage is that they
are naturally broken up into smaller time units by quarterly (three month)
breaks.
Splitting the data set into quarters, rather than modelling the entire time
series contiguously speeds up the computation time as inverting several
small matrices is faster than inverting one large one.
The \kepler\ quarter divisions are natural places to split the time series
because the spacecraft rotates by ninety degrees every quarter (three months),
placing each star on a new CCD module.
Pixel response functions and background flux differs from pixel-to-pixel and
module-to-module, so noise properties of \kepler\ light curves change every
quarter.
Additionally, changes in the spacecraft's orientation and position during
quarterly re-pointings temporarily affect the temperature of the CCD,
producing systematic features in the light curves at the start of some
quarters.
We model each quarter separately but the parameters of the GP kernel function
are global, {\it i.e.} we do not use a separate period parameter for each
quarter---there is just one period parameter for an entire light curve.
It would be possible to model the time series with a mixture of some global
parameters and some quarter-specific parameters, for example one might expect
that the amplitude of covariance, $A$ or white noise level, $\sigma$ to vary
on a quarterly basis.
However, since there are seventeen quarters this would lead to thirty-seven
parameters, and in the interest of minimising computation time (adding
parameters leads to longer MCMC burn in and convergence time), we choose to
use global parameters only.
Unfortunately, the application of this method to noisy test cases is still
under development and cannot provide results here.
We hope to develop this method further and apply it to real light curves in
the near future.

% The marginal posterior distributions of the QP kernel hyperparameters for an
% example noisy light curve  with a ... day rotation period, are shown in
% figure \ref{fig:gp_posteriors}.
% The light curve itself with the best fit GP model is shown in figure
% \ref{fig:demo_lc_GP}.

% \begin{figure*}
% \begin{center}
% \includegraphics[width=6in, clip=true]{figures/demo_lc_GP.pdf}
% \caption{A simulated light curve with a rotation period of 14.4 days.
% The pink curve shows the GP model with the best-fit parameters.
% The marginal posteriors of the GP hyper-parameters used to produce this fit
% are shown in figure \ref{fig:gp_posteriors}.}
% \label{fig:demo_lc_GP}
% \end{center}
% \end{figure*}

% \begin{figure*}
% \begin{center}
% \includegraphics[width=6in, clip=true]{figures/demo_triangle.pdf}
% \caption{Marginal posterior PDFs of the QP GP model parameters. $\sigma$ is an
% additional white noise term added to the diagonal elements of the covariance
% matrix. It is the fraction by which the observational uncertainties have been
% underestimated. If the errorbars reported on the data are too small, this
% parameter will be non-zero. Since this light curve was simulated without
% noise, this parameter is very close to zero.
% The true rotation period of this star is 14.4 days.}
% \label{fig:gp_posteriors}
% \end{center}
% \end{figure*}

@@

The main drawback of the GP method is computation time.
Because GPs are so expensive to compute, it is necessary to come up with
shortcuts in order to perform inference on \kepler\ light curves, each of
which compromises accuracy.
The shortcuts that we employ here include:
\begin{itemize}
\item{Initialisation.
The choice of initial parameter values makes a difference to computation time:
the closer they are to the maximum likelihood parameters the shorter the MCMC
burn-in time.
Using the ACF period to initialise is not ideal as results will not be fully
independent of the ACF period unless MCMC chains are run for an infinitely
long time.} \item{Subsampling.
Not only is some information always lost by subsampling, but the choice of
subsampling frequency will also effect the results.
We chose to retain only 20 data points per period --- based on the ACF
period --- which, again means that our results are not independent of the ACF
results.}
\item{Priors.
We chose to use uniform, bounded priors to reduce parameter space and
therefore convergence time.
Unfortunately, if the true period lies outside the bounded region, it will
never be found.}
\end{itemize}

These shortcuts are only necessary when working with a large number of stars
(of order hundreds or thousands).
If only interested in single or small numbers of targets it may be possible to
reduce the extremity of these shortcuts or to avoid them altogether.

\subsection{The ACF method}
The main advantage of the ACF method over the GP method is speed: it is
extremely fast and therefore useful to get a quick estimate of a period.
We have demonstrated that the ACF method may produce rotation periods that are
slightly systematically biased towards lower rotation periods.
This effect comes from the process of extracting the period from the ACF
itself: typically superimposed onto a decaying exponential, that first peak in
the ACF gets shifted towards faster periods.
One may be able to get around this by modelling the ACF as a sum of a cosine
and exponential function (a similar function to GP kernel function) but
in practice ACFs tend to have a more complicated structure and it is not
possible to model them with a simple function.
This observation leads to the idea of modelling the ACF with a Gaussian
processes, which then logically progresses to a using a Gaussian process with
a Gaussian process kernel function.
Although this may be an attractive idea, in practice a GP would not
necessarily produce a positive semi-definite covariance kernel\footnote{One
that can be inverted; a necessary regression operation.}.
An alternative approach may be to correct for rotation period bias after the
peak position has been measured by inferring a correction factor.
Yet another approach could be to measure the mean separation between adjacent
peaks in the ACF: the first peak should be most strongly affected by the
exponential decay, the second peak less so, and so on.

\subsection{Initialisation}
Using the ACF to initialise the MCMC chains is not ideal because, of course,
one becomes reliant on the assumption that the ACF period is close to the true
period.
Of course, if you had infinite CPU time, this would not be a problem as you
would eventually sample the entire posterior PDF of the period parameter,
however in practice this is likely to be an issue.
The only way to get around this problem is to run the MCMC chains for as long
as possible, or, alternatively to use a sampler that is designed to move
around the parameter space much more quickly than {\tt emcee}; nested sampling
for example.
We chose to use the ACF periods rather than the periodogram periods to
initialise as, although there was some systematic bias present in these
results, that bias was small and there were fewer large outliers.

>>>>

\newcommand{\columbia}{1}
\newcommand{\oxford}{2}
\newcommand{\uw}{3}
\newcommand{\princeton}{4}
\newcommand{\naigrain}{333}
\newcommand{\nmcquillan}{100}
\newcommand{\Kepler}{{\it Kepler}}
\newcommand{\kepler}{\Kepler}
\newcommand{\corot}{{\it CoRoT}}
\newcommand{\Ktwo}{{\it K2}}
\newcommand{\ktwo}{\Ktwo}
\newcommand{\TESS}{{\it TESS}}
\newcommand{\LSST}{{\it LSST}}
\newcommand{\SDSS}{{\it SDSS}}
\newcommand{\PLATO}{{\it PLATO}}
\newcommand{\Teff}{$T_{\mathrm{eff}}$}
\newcommand{\teff}{$T_{\mathrm{eff}}$}
\newcommand{\FeH}{[Fe/H]}
\newcommand{\feh}{[Fe/H]}
\newcommand{\ie}{{\it i.e.}}
\newcommand{\eg}{{\it e.g.}}
\newcommand{\logg}{log \emph{g}}
\newcommand{\dnu}{$\Delta \nu$}
\newcommand{\numax}{$\nu_{\mathrm{max}}$}
\newcommand{\acfRMS}{1.45}
\newcommand{\pgramRMS}{1.10}
\newcommand{\mcmcRMS}{0.87}

@@

\altaffiltext{\columbia}{Simons Fellow, Department of Astronomy, Columbia
University, NY, NY, RuthAngus@gmail.com}
\altaffiltext{\oxford}{Subdepartment of Astrophysics, University of Oxford,
UK}
\altaffiltext{\uw}{Department of Astronomy, University of Washington, Seattle,
WA}
\altaffiltext{\princeton}{Department of Astronomy, Princeton University,
Princeton, NJ}

\begin{abstract}
The light curves of spotted, rotating stars are often non-sinusoidal and
quasi-periodic---spots are not static on the stellar surface and have finite
lifetimes causing stellar flux variations to slowly shift in phase.
A strictly periodic sinusoid is therefore not a representative model for the
light curve of star with a rotational signal.
Ideally, a physical model of the stellar surface would be conditioned on the
data, however the parameters of such models can be highly degenerate
\citep[\eg][]{Russell1906, Jeffers2009}.
Instead, we use an appropriate {\it effective} model: a Gaussian Process (GP)
with a quasi-periodic covariance kernel function.
By modelling the light curve with a GP, a highly flexible semi-parametric
function, we avoid the necessity of choosing a `best fitting' functional form,
whilst sampling directly from the posterior Probability Density Function (PDF)
of the periodic parameter and marginalising over the other kernel
hyperparameters.
We used \naigrain\ simulated light curves with a range of rotation periods
to test the GP model.
We attempted to recover the rotation periods using three methods: our GP
method, a sine-fitting periodogram method and an AutoCorrelation Function
(ACF) method.
The posterior PDF of the rotation period parameter was sampled using the
affine invariant ensemble Markov Chain Monte Carlo (MCMC) sampler {\tt emcee}
\citep{Foreman-Mackey2013}, and the GP operations were performed using the
{\tt george} python package \citep{George}.
Rotation periods inferred via this method are more precise and accurate than
the periodogram and ACF methods.
Furthermore, rotation periods are inferred probabilistically, allowing them to
be easily incorporated into follow-on studies, inferring stellar ages or
characterising star-planet interactions for example.
In addition, unlike the sine-fitting periodogram and ACF methods, the
posterior PDF samples allow for proper characterisation of uncertainties.
This method eliminates the need to define a heuristic detection criterion such
as are often used in the literature: a light curve that does not contain a
rotation signal will return the prior PDF.

% Furthermore, the improvement is expected to be even more dramatic when applied
% to real, noisy {\it Kepler} light curves, since the GP method is well suited
% to modelling rotation signals and correlated noise simultaneously.

@@

In order to test the GP method we attempted to the measure rotation periods of
a set of simulated light curves.
We used \naigrain\ light curves generated for the \citet{Aigrain2015}
`hare and hounds' rotation period recovery experiment.
These light curves were simulated by placing dark, circular spots with slowly
evolving size, on the surface of bright, rotating spheres, ignoring
limb-darkening effects.
\citet{Aigrain2015} simulated one thousand light curves in order to test the
ability of participating teams to recover both the stellar rotation periods
and the rotational shear: the amplitude of surface differential rotation.
% Since we are not interested in recovering differential rotation we only used
% those light curves simulated {\it without} differential rotation, of which
% there are \naigraincurves.
For the purpose of demonstrating that our method can recover rotation periods,
we selected only the \naigrain\ out of the 1000 without differential
rotation.
We opted to use only solid-body rotators because differential rotation may
produce some additional scatter in the rotation period measurements recovered.
In future we intend to test whether we can recover differential rotation using
the GP method and will then use the full set of 1000 light curves.
The distributions of input parameters used to generate these light curves are
listed in \citet{Aigrain2015}.
Light curves were simulated with a real \Kepler\ long-cadence time array:
one data point every thirty minutes over a four year duration.
90\% of the rotation periods of the simulations were randomly drawn from a
log-uniform distribution between 10 and 50 days and 10\% from a log-uniform
distribution between 1 and 10 days.
The light curves were also generated with a range of stellar inclination
angles, activity levels, spot lifetimes and more.
Once simulated, the light curves were added to real \kepler\ light curves with
no obvious astrophysical variability in order to preserve the noise properties
of the time series.
Figure \ref{fig:demo_lc} shows an example of a simulated, noise-free
light curve with a period of 17.04 days.

@@

\begin{table*}
\begin{center}
\caption{Ranges and distributions of parameters used to simulate light curves
in \citet{Aigrain2015}}
\begin{tabular}{lcc}
\hline\hline
    Parameter & Range & Distribution \\
    \hline
    Rotation period, $P_{rot}$ & 10 - 50 days (90\%) & log uniform \\
    & 1 - 10 days (10\%) & log uniform \\
    Activity cycle length & 1 - 10 years & log uniform \\
    Inclination & 0 - 90$^\circ$ & Uniform in $\sin^2i$ \\
    Decay timescale & (1 - 10) $\times P_{rot}$ & log uniform \\
\hline
\end{tabular}
\end{center}
\end{table*}
\label{tab:simulation_parameters}

@@

We attempted to recover the rotation periods of these \naigrain\ light
curves using three rotation period recovery methods: the ACF method, the LS
periodogram method and the GP method.

@@

We applied the ACF method to the sample of \naigrain\ simulated light
curves.
Periods measured using the ACF method are plotted against the original
rotation period values used to generate the light curves in figure
\ref{fig:compare_acf}.
The $2\times$ and $1/2\times$ harmonic lines are shown as dashed lines.
The RMS scatter about the $x=y$ line for rotation periods measured using the
ACF method is \acfRMS.
The agreement between the injected and recovered rotation periods is good in
general, however twice the true rotation period is often found.
In addition, some rotation periods are drastically over- and underestimated.
Of the points clustered close to the $x=y$ line, more slightly below it than
above it, \ie\ rotation periods are preferentially underestimated.
This is a feature of the peak position measurement method that is performed on
the ACF.
ACFs of stellar light curves are, like the QP kernel function in equation
\ref{eq:QP}, periodic functions added to decaying exponentials.
In such functions the peak positions can be shifted towards the left, the
short period end, because the decaying exponential raises the left side of
each peak more than the right.
It is possible to model this effect, however the standard practice is to
simply measure the peak position without taking it into account.
We adopt the standard practice approach here in order to provide a comparison
between our new method and the current standard procedure.

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/compare_acf.pdf}
\caption[ACF results.]
{The `true' rotation periods used to generate \naigrain\ simulated light
curves vs the rotation periods measured using the ACF technique.}
\label{fig:compare_acf}
\end{center}
\end{figure*}

@@

For each simulated light curve, a Lomb-Scargle (LS) periodogram
\citep{Lomb1976, Scargle1982} was computed over a grid of 10,000 periods,
evenly spaced in frequency, between 1 and 100 days\footnote{LS periodograms
were calculated using the gatspy Python module:
\url{https://github.com/astroML/gatspy/tree/master/gatspy/periodic}.}
The period of the highest peak in the periodogram was adopted as the rotation
period.
The resulting recovered rotation periods are plotted as a function of the true
periods in figure \ref{fig:pgram_compare}.
These recovered rotation periods are in general more accurate than the ACF
results: they do not systematically over or under predict rotation period.
The RMS scatter for rotation periods measured using the LS periodogram
technique is \pgramRMS.
The longer-period measurements provide the dominant contribution to this RMS.
% We believe that the unsuccessful recoveries occurring preferentially for
% shorter rotation periods are produced by additional, longer timescale
% variations in some of the light curves.
% Specifically, by fluctuations produced by changes in the total spot coverage
% on the stellar surface.
% These fluctuations can occasionally be larger in magnitude than those produced
% by the rotational signal itself and are present in the majority of outlying
% light curves that occupy the top left of figure
% \ref{fig:pgram_compare_noise-free}.

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/compare_pgram.pdf}
\caption[LS periodogram results.]
{The `true' rotation periods used to generate \naigrain\ simulated light
curves vs the rotation periods measured using a LS periodogram technique.}
\label{fig:pgram_compare}
\end{center}
\end{figure*}

@@

\begin{table*}
\begin{center}
\caption{Means and standard deviations of the Gaussian priors on the natural
logarithms of the GP model parameters.}
\begin{tabular}{lcc}
Parameter & Mean & Standard Deviation \\
    \hline
    $\ln(A)$ & -13 & 2.7 \\
    $\ln(l)$ & 6.2 & 1.5 \\
    $\ln(\Gamma)$ & -1.4 & 1.5 \\
    $\ln$(Rotation period, $P_{rot}$) & $\ln$(ACF period) &
                2 $\times \ln$(ACF period) \\
    $\ln\sigma$ & -17 & 5 \\
\end{tabular}
\end{center}
\end{table*}
\label{tab:priors}
The parameter positions were also initialised at the means of these
distributions during MCMC posterior exploration.
In addition, we set the following bounds on the rotation period and the $l$
parameter.
\begin{eqnarray}
 	&	\ln(0.5) < \ln(P_{\mathrm{rot}}) < \ln(T/2) \\ \nonumber
    &   0 < \ln(l) \\
\end{eqnarray}
\label{eq:priors}
Where $T$ is the total timespan of the light curve.
We limit rotation periods to be greater than half a day because we are chiefly
interested in rotation of FGK stars which are likely to be greater than this
lower limit, however this restriction could be relaxed in principle.
We also limit the rotation periods to be less than half of the total timespan
of the data since this is the longest period we expect to reliably recover.
We prevent $\ln(l)$ from falling beneath zero both because we expect that
astrophysical signals in light curves should have longer correlation
timescales than this and in practice we found that if $l$ is unrestricted, it
can occasionally wander towards a small value, resulting in an incorrect
rotation period estimate.

@@

We subsampled the light curves in order to reduce computation time.
We randomly subsampled the data, retaining an average of four data points per
day.
This allowed us to recover periods down to half a day --- the minimum
timescale of interest for this project, whilst reducing computational costs.
In order to reduce computation time the light curves were split into small
sections and the likelihood for each section calculated individually, then
summed.
The matrix operations scale as $N\ln(N)^2$ where $N$ is the number of data
points in the light curve, however the likelihood calculation scales linearly
with the number of light curve sections used.
By adding together likelihoods of individial sections, the likelihood
evaluation scales as $mn\ln(n)^2$, where $n$ is the number of data points per
section and $m$ is the number of sections.
This method therefore significantly reduces computation time.
We split the simulated light curves into sections, 200 days long.
The {\tt george} \citep{George} python package was used to implement the GP
model which uses the fast matrix solver, HODLR \citep{Ambikasaran2014}.
Matrix operations performed with HODLR scale as $N \log ^2 (N)$, where $N$ is
the number of data points.
We used {\tt emcee} \citep{Foreman-Mackey2013}, an affine invariant
\citep{Goodman2010} ensemble sampler to explore the posterior PDFs of the
model parameters.
The rotation period was taken as the sample with the maximum likelihood value.
The resulting measured rotation periods are compared to the true rotation
periods in figure \ref{fig:compare_mcmc}.

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/compare_mcmc.pdf}
\caption{The `true' rotation periods used to generate \naigrain\
simulated light curves vs the rotation periods measured using the ACF
technique.}
\label{fig:compare_mcmc}
\end{center}
\end{figure*}

The marginal posterior distributions of the QP kernel hyperparameters for the
example simulated light curve with in figure \ref{fig:demo_lc}, are shown in
figure \ref{fig:gp_posteriors}.

\begin{figure*}
\begin{center}
\includegraphics[width=6in, clip=true]{figures/0002_triangle.png}
\caption{(Place holder - to be updated.
Marginal posterior PDFs of the QP GP model parameters. $\sigma$ is an
additional white noise term added to the diagonal elements of the covariance
matrix. It is the fraction by which the observational uncertainties have been
underestimated. If the uncertainties reported on the data are too small, this
parameter will be non-zero.
The true rotation period of this star is ... ($\exp(...)$) days.}
\label{fig:gp_posteriors}
\end{center}
\end{figure*}

@@

We applied our method to \nmcquillan\ stars with rotation periods previously
measured by \citet{Mcquillan2014}.
A feature of \kepler\ light curves that works to our advantage is that they
are naturally broken up into smaller time units by quarterly (three month)
breaks.
As explained above, splitting the data set into quarters, rather than
modelling the entire time series contiguously speeds up the computation time
as inverting several small matrices is faster than inverting one large one.
The \kepler\ quarter divisions are natural places to split the time series
because the spacecraft rotates by ninety degrees every quarter (three months),
placing each star on a new CCD module.
Pixel response functions and background flux differs from pixel-to-pixel and
module-to-module, so noise properties of \kepler\ light curves change every
quarter.
Additionally, changes in the spacecraft's orientation and position during
quarterly re-pointings temporarily affect the temperature of the CCD,
producing systematic features in the light curves at the start of some
quarters.
We model each quarter separately but the parameters of the GP kernel function
are global, {\it i.e.} we do not use a separate period parameter for each
quarter---there is just one period parameter for an entire light curve.
It would be possible to model the time series with a mixture of some global
parameters and some quarter-specific parameters, for example one might expect
that the amplitude of covariance, $A$ or white noise level, $\sigma$ to vary
on a quarterly basis.
However, since there are seventeen quarters this would lead to thirty-seven
parameters, and in the interest of minimising computation time (adding
parameters leads to longer MCMC burn in and convergence time), we choose to
use global parameters only.

Our results show good agreement with the \citet{Mcquillan2014} measurements,
as shown in ... %\ref{fig:mcquillan}.

@@

The main drawback of the GP method is computation time.
Because GPs are so expensive to compute, it is necessary to come up with
shortcuts in order to perform inference on \kepler\ light curves, each of
which compromises accuracy.
These shortcuts may only be necessary when working with a large number of
stars (of order hundreds or thousands).
If only interested in single or small numbers of targets it may be possible to
reduce the extremity of these shortcuts or to avoid them altogether.
The shortcuts that we employ here include:
\begin{itemize}
\item{Initialisation.
The choice of initial parameter values makes a difference to computation time:
the closer they are to the maximum likelihood parameters the shorter the MCMC
burn-in time.
Using the ACF period to initialise is not ideal as results will not be fully
independent of the ACF period unless MCMC chains are run for an infinitely
long time.}
\item{Subsampling.
Not only is some information always lost by subsampling, but the choice of
subsampling frequency will also effect the results.
We choose to retain an average of only 4 data points per day, based on the
range of rotation periods we are interested in.}
\item{Priors.
We choose to place a Gaussian prior in log-space over the rotation period
parameter where the standard deviation is twice the log of the ACF rotation
period.
Unfortunately however, if the true period is far different to the ACF period,
the MCMC chains will take a long time to reach the correct value.
In practice we found that the introduction of a Gaussian prior improved
results compared with a broad prior that was uniform in log-space.}
\end{itemize}

\subsection{Initialisation}
Using the ACF to initialise the MCMC chains is not ideal because, of course,
one becomes reliant on the assumption that the ACF period is close to the true
period.
Of course, if you had infinite CPU time, this would not be a problem as you
would eventually sample the entire posterior PDF of the period parameter,
however in practice this is likely to be an issue.
The only way to get around this problem is to run the MCMC chains for as long
as possible, or, alternatively to use a sampler that is designed to move
around the parameter space much more quickly than {\tt emcee}.
% We implemented a version of our code using nested sampling but the results
% were not as good as the {\tt emcee} method.
% We believe this to be due to the initialisation --- there is no initialisation
% in nested sampling, samples are drawn directly from the prior.
% In a problem with a multimodal posterior PDF such as inferring periods,
% initialising MCMC chains close to the maximum likelihood value may lead to a
% more rapid convergence.
We chose to use the ACF periods rather than the periodogram periods to
initialise the results since there were fewer large outliers.
